---
title: "SpatialR"
author: "Moegamad Uzair Jack"
date: "2025-02-24"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
   echo = TRUE,      # Show code
  message = FALSE,  # Hide package loading messages
  warning = FALSE  # Hide warnings
)

```

## Loading packages
Loading all necessary packages
```{r}
# Load libraries
library(rgbif)      # Access GBIF data
library(sf)         # Handle spatial data
library(ggplot2)    # Create the map
library(dplyr)      # Data wrangling
library(viridis)    # Colour scales
library(leaflet)
library(rnaturalearth)
```
## Get Nightjar Data from GBIF
We’ll query GBIF for Caprimulgidae sightings, limited to 1000 records for simplicity.
```{r}
# Get the GBIF species key for the Nightjar family (Caprimulgidae)
nightjar_family_key <- name_backbone(name = "Caprimulgidae")$familyKey

# Search for occurrence data (optimized query)
nightjar_data <- occ_search(
  taxonKey = nightjar_family_key,
  hasCoordinate = TRUE,
  basisOfRecord = "HUMAN_OBSERVATION",
  limit = 5000
)
```

## Clean and Prepare Data
Select the important columns and remove any rows missing coordinates
```{r}
# Clean and prepare the data
nightjar_sightings <- nightjar_data$data %>%
  select(species, decimalLatitude, decimalLongitude, eventDate, country, 
         individualCount) %>%
  na.omit()  # Removes rows with any NA values
```
## Convert to Spatial Data
Turn this into an sf (Simple Features) object so it’s ready for mapping.
```{r}
nightjar_sf <- st_as_sf(nightjar_sightings,
                        coords = c("decimalLongitude", "decimalLatitude"),
                        crs = 4326)  # WGS84 projection


```
## Create  static map
```{r}
# Get a world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Create the map
ggplot(data = world) +
  geom_sf(color = "gray70") +
  geom_point(data = nightjar_sf, 
             aes(x = st_coordinates(nightjar_sf)[,1], 
                 y = st_coordinates(nightjar_sf)[,2], 
                 color = factor(species), 
                 size = individualCount),
             alpha = 0.7) +
  scale_color_viridis_d(option = "plasma") +
  theme_minimal() +
  labs(title = "Global Nightjar Sightings",
       x = "Longitude",
       y = "Latitude",
       color = "Species",
       size = "Count")

```

## CRS corrections
```{r}
st_crs(nightjar_sf)

```


## Interactive Nightjar Sightings Map
Create an interactive map where you can zoom, pan, and click on individual sightings to see details.
Adding a Custom base map with species specific colour markers

```{r}
# Ensure individualCount is numeric and remove any NAs just in case
nightjar_sf$individualCount <- as.numeric(nightjar_sf$individualCount)
nightjar_sf <- nightjar_sf %>%
  filter(!is.na(individualCount))

# Create a color palette for species (categorical data)
species_palette <- colorFactor(palette = "plasma", 
                               domain = nightjar_sf$species)

# Map with species-based color function
leaflet(nightjar_sf) %>%
  addProviderTiles(providers$Esri.WorldPhysical) %>%
  addCircleMarkers(
    radius = 3,  # Fixed size 
    color = ~species_palette(species),  # Different color for each species
    fillOpacity = 0.8,
    popup = ~paste0(     # Creating popup with relevant infornmation
      "<strong>Species: </strong>", species,
      "<br><strong>Count: </strong>", individualCount,
      "<br><strong>Country: </strong>", country,
      "<br><strong>Date: </strong>", eventDate
    )
  ) %>%
  setView(lng = 20, lat = 0, zoom = 3) %>% # Centre  map on Africa
  
  addScaleBar ( position = "topright" )

```





